---
title: "Data_Analysis"
output: html_document
date: "2026-06-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
library(lmerTest)
library(lme4)
library(tidyr)
library(ggplot2)
library(ggnewscale)
library(dplyr)
library(tidyr)
library(broom.mixed)
library(tidyr)
library(purrr)
library(tidyverse)
library(jsonlite)
library(nnet)
library(patchwork)
library(emmeans)

```


## Import Data
<!-- Load individual data -->

```{r, include = FALSE}
df_1339 <- read.delim("Data/TwoStep_CCNP001339_1_13-Mar-2025.txt")

df_1445 <- read.delim("Data/TwoStep_CCNP001445_1_12-Apr-2025.txt")

df_1604 <- read.delim("Data/TwoStep_CCNP001604_1_13-Mar-2025.txt")

df_1613 <- read.delim("Data/TwoStep_CCNP001613_1_05-Mar-2025.txt")

df_1638 <- read.delim("Data/TwoStep_CCNP001638_1_18-Mar-2025.txt")

df_1665 <- read.delim("Data/TwoStep_CCNP001665_1_12-Apr-2025.txt")

df_1670 <- read.delim("Data/TwoStep_CCNP001670_1_25-Apr-2025.txt")

df_1674 <- read.delim("Data/TwoStep_CCNP001674_1_23-Apr-2025.txt")

df_1678 <- read.delim("Data/TwoStep_CCNP001678_1_30-Apr-2025.txt")

df_1679 <- read.delim("Data/TwoStep_CCNP001679_1_28-Apr-2025.txt")

df_1684 <- read.delim("Data/TwoStep_CCNP001684_1_01-May-2025.txt")

df_1689 <- read.delim("Data/TwoStep_CCNP001689_1_30-Apr-2025.txt")

patient_data_text <- read.csv("Data/Gueguen_01_Subject+Questionnaire+-+Updated_June+24,+2026_12.54.csv")

patient_data_text <- patient_data_text[-c(1, 2), ]

patient_data_num <- read.csv ("Data/Gueguen_01_Subject+Questionnaire+-+Updated_June+24,+2026_12.43.csv") 

patient_data_num <- patient_data_num[-c(1, 2, 3), ]


```

<!-- Combine all data: import CSV -->

```{r, include = FALSE}

# Create output directory
output_dir <- "data_r"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Location of raw txt files
data_dir <- "Data"

# Find all TwoStep files
files <- sort(list.files(
  path = data_dir,
  pattern = "^TwoStep_CCNP.*\\.txt$",
  full.names = TRUE
))

# Create manifest
manifest <- tibble(path = files) %>%
  mutate(
    source_file = basename(path),
    participant_id = str_extract(source_file, "CCNP\\d+")
  )

# Save participant IDs
participant_ids <- manifest$participant_id
saveRDS(participant_ids,
        file.path(output_dir, "participant_ids.RDS"))

# Read and combine all files
df_all <- manifest %>%
  mutate(
    data = map(path, ~ read.delim(
      .x,
      stringsAsFactors = FALSE,
      colClasses = "character"
    ))
  ) %>%
  select(-path) %>%
  unnest(data)

# Save combined dataset
saveRDS(df_all, file.path(output_dir, "df_all.RDS"))
```


## Data Cleaning

```{r, echo = FALSE}

df_all <- df_all %>%
  mutate(
    trial   = as.numeric(trial),
    choice1 = as.numeric(choice1),
    choice2 = as.numeric(choice2),
    won     = as.numeric(won),
    state   = as.numeric(state),

    invalid_trial = !(choice1 %in% c(1,2) &
                      choice2 %in% c(1,2))
  )
```

```{r, echo = FALSE}
stay_df <- df_all %>%
  arrange(participant_id, trial) %>%
  group_by(participant_id) %>%
  mutate(
    transition = case_when(
      choice1 == 1 & state == 2 ~ "Common",
      choice1 == 2 & state == 3 ~ "Common",
      TRUE ~ "Rare"
    ),

    # was previous trial invalid?
    # if True: means the first trial 
    prev_invalid = lag(invalid_trial),

    stay = case_when(
      is.na(prev_invalid) ~ NA_real_,      # first trial
      prev_invalid ~ NA_real_,             # previous trial invalid
      TRUE ~ as.numeric(choice1 == lag(choice1))
    ),

    prev_reward = case_when(
      is.na(prev_invalid) ~ NA,
      prev_invalid ~ NA,
      TRUE ~ lag(won)
    ),

    prev_transition = case_when(
      is.na(prev_invalid) ~ NA_character_,
      prev_invalid ~ NA_character_,
      TRUE ~ lag(transition)
    )
  ) %>%
  ungroup()

stay_df

```

### Sanity Check

```{r, echo = FALSE}
df_all <- df_all %>%
  mutate(
    invalid_trial = !(choice1 %in% c(1,2) &
                      choice2 %in% c(1,2))
  )

# total invalid trials = 43
sum(df_all$invalid_trial)

# by subject
df_all %>%
  filter(invalid_trial) %>%
  count(participant_id)

# df_all %>%
#   filter(invalid_trial) %>%
#   select(participant_id, trial, choice1, choice2)
```

There are 55 lag_missing trials with NA:

- 12 first trials
- 30 missing choice_1
- 13 missing choice_2

```{r, echo = FALSE}
stay_df <- stay_df %>%
  mutate(
    lag_missing = is.na(stay)
  )

lag_missing_trials <- stay_df %>%
  filter(lag_missing)

lag_missing_trials
```

```{r, echo = FALSE}
# remove the invalid 43 trials
stay_df_clean <- stay_df %>%
  filter(
    choice1 %in% c(1, 2),
    choice2 %in% c(1, 2)
  )
```


## Proportion of Choices

```{r, echo = FALSE}
# Group Level proportion of choice 1
stay_df_clean %>%
  count(choice1) %>%
  mutate(prop = n / sum(n))

# Group Level proportion of choice 2
stay_df_clean %>%
  count(choice2) %>%
  mutate(prop = n / sum(n))

```

```{r, echo = FALSE}
# Ind level proportion of choice 1
choice1_prop_subj <- stay_df_clean %>%
  group_by(participant_id) %>%
  summarise(
    prop_choice1_1 = mean(choice1 == 1),
    prop_choice1_2 = mean(choice1 == 2),
    n_trials = n()
  )

choice1_prop_subj

# Ind level proportion of choice 2
choice2_prop_subj <- stay_df_clean %>%
  group_by(participant_id) %>%
  summarise(
    prop_choice2_1 = mean(choice2 == 1),
    prop_choice2_2 = mean(choice2 == 2),
    n_trials = n()
  )

choice2_prop_subj
```

```{r, echo = FALSE}
ggplot(choice1_prop_subj,
       aes(x = factor(substr(participant_ids, nchar(participant_ids) - 3, nchar(participant_ids))),
           y = prop_choice1_1)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = .5,
             linetype = "dashed") +
  ylim(0, 1) +
  labs(
    x = "Subject",
    y = "Proportion Choice1 = 1"
  )
```
```{r, echo = FALSE}
ggplot(choice2_prop_subj,
       aes(x = factor(substr(participant_ids, nchar(participant_ids) - 3, nchar(participant_ids))),
           y = prop_choice2_1)) +
  geom_col(fill = "purple") +
  geom_hline(yintercept = .5,
             linetype = "dashed") +
  ylim(0, 1) +
  labs(
    x = "Subject",
    y = "Proportion Choice2 = 1"
  )
```

### Summary 

```{r, echo = FALSE}
choice1_prop_subj %>%
  summarise(
    mean_prop_choice1_1 = mean(prop_choice1_1),
    sd_prop_choice1_1   = sd(prop_choice1_1),
    min_prop_choice1_1  = min(prop_choice1_1),
    max_prop_choice1_1  = max(prop_choice1_1)
  )

choice2_prop_subj %>%
  summarise(
    mean_prop_choice2_1 = mean(prop_choice2_1),
    sd_prop_choice2_1   = sd(prop_choice2_1),
    min_prop_choice2_1  = min(prop_choice2_1),
    max_prop_choice2_1  = max(prop_choice2_1)
  )
```

## Stay Probability vs Common/Rare

choice 1 --> state 2 is common ~69%

choice 1 --> state 3 is rare ~31%

choice 2 --> state 3 is common ~69%

choice 2 --> state 2 is rare ~32%

```{r, echo = FALSE}
# check the transition probability
stay_df_clean %>%
  count(choice1, state) %>%
  group_by(choice1) %>%
  mutate(prop = n / sum(n))

# Define common & rare:
stay_df_clean <- stay_df_clean %>%
  mutate(
    transition = case_when(
      choice1 == 1 & state == 2 ~ "Common",
      choice1 == 2 & state == 3 ~ "Common",
      choice1 == 1 & state == 3 ~ "Rare",
      choice1 == 2 & state == 2 ~ "Rare",
      TRUE ~ NA_character_
    )
  )
```

```{r}

# remove all the trials with NA (e.g. all the first trials where there is no lagging)

stay_analysis <- stay_df_clean %>%
  filter(
    !is.na(stay),
    !is.na(prev_reward),
    !is.na(prev_transition)
  )

stay_summary <- stay_analysis %>%
  mutate(
    prev_reward_label = ifelse(prev_reward == 1,
                               "Rewarded",
                               "Unrewarded")
  ) %>%
  group_by(participant_id,
           prev_reward_label,
           prev_transition) %>%
  summarise(
    stay_prob = mean(stay, na.rm = TRUE),
    .groups = "drop"
  )

stay_summary
```

```{r}
stay_summary <- stay_analysis %>%
  mutate(prev_reward_label = ifelse(prev_reward == 1, "Rewarded", "Unrewarded")) %>%
  group_by(participant_id, prev_reward_label, prev_transition) %>%
  summarise(
    stay_prob = mean(stay),
    se = sd(stay) / sqrt(n()),
    .groups = "drop"
  )
```

```{r}
#compute the stay probability & lag the reward & transition

stay_df <- stay_df %>%
  arrange(participant_id, trial) %>%
  group_by(participant_id) %>%
  mutate(
    # Stay: repeated same choice1 as previous trial
    stay = as.integer(choice1 == lag(choice1)),
    
    # Previous trial's reward and transition 
    prev_reward = lag(won),
    prev_transition = lag(transition)
  ) %>%
  ungroup() %>%
  # Remove first trial of each subject 
  filter(!is.na(stay), !is.na(prev_reward), !is.na(prev_transition))

stay_summary <- stay_df %>%
  mutate(prev_reward_label = ifelse(prev_reward == 1, "Rewarded", "Unrewarded")) %>%
  group_by(participant_id, prev_reward_label, prev_transition) %>%
  summarise(stay_prob = mean(stay), .groups = "drop")

head(stay_summary)
```

```{r}
stay_summary <- stay_df %>%
  mutate(prev_reward_label = ifelse(prev_reward == 1, "Rewarded", "Unrewarded")) %>%
  group_by(participant_id, prev_reward_label, prev_transition) %>%
  summarise(
    stay_prob = mean(stay),
    se = sd(stay) / sqrt(n()),  # standard error
    .groups = "drop"
  )
```

### Plot for each subj

```{r, echo = FALSE}
ggplot(stay_summary, aes(x = prev_reward_label, y = stay_prob, fill = prev_transition)) +
  geom_col(position = position_dodge(width = 0.7),
           color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = stay_prob - se, ymax = stay_prob + se),
                position = position_dodge(width = 0.7),
                width = 0.2) +
  scale_fill_manual(values = c("Common" = "#cde5f9", "Rare" = "#ebbdc8"),
                    name = "Transition") +
  scale_x_discrete(limits = c("Rewarded", "Unrewarded")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Previous Trial",
    y = "Stay Probability"
  ) +
  theme_classic() +
  theme(legend.position = "bottom") +
  facet_wrap(~ participant_id, nrow = 4, ncol = 3)

ggsave("stay_probability.png", width = 8, height = 6, dpi = 300)
```

```{r, include = FALSE}
# Example for each ind. plot
stay_summary %>%
  filter(participant_id == "CCNP001339") %>%
  ggplot(aes(x = prev_reward_label, y = stay_prob, fill = prev_transition)) +
  geom_col(position = position_dodge(width = 0.7),
           color = "black", width = 0.6) +
  scale_fill_manual(values = c("Common" = "#cde5f9", "Rare" = "#ebbdc8"),
                    name = "Transition") +
  scale_x_discrete(limits = c("Rewarded", "Unrewarded")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Subject:CCNP001339",
    x = "Previous Trial",
    y = "Stay Probability"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## Check Transition Probability 

```{r}
# 1. Per rocket transition probability
rocket_transition <- stay_df_clean %>%
  group_by(choice1, state) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(choice1) %>%
  mutate(prop = n / sum(n),
         transition = case_when(
           choice1 == 1 & state == 2 ~ "Common",
           choice1 == 2 & state == 3 ~ "Common",
           TRUE ~ "Rare"
         ))

print(rocket_transition)

# 2. Overall transition probability (collapsed across rockets)
overall_transition <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(transition) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(prop = n / sum(n))

print(overall_transition)

# 3. Block transition fraction (overall)
block_transition <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(participant_id, transition) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(participant_id) %>%
  mutate(prop = n / sum(n))

print(block_transition)

# 4. Rocket transition fraction (by rocket, per subject)
rocket_transition_subj <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(participant_id, choice1, transition) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(participant_id, choice1) %>%
  mutate(prop = n / sum(n))

print(rocket_transition_subj)
```

## Data visualization for transition probability

```{r}
# overall_transition with SE across subjects
overall_transition <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(participant_id, transition) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(participant_id) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  group_by(transition) %>%
  summarise(
    mean_prop = mean(prop),
    se = sd(prop) / sqrt(n()),
    .groups = "drop"
  )
```

```{r, echo = FALSE}
ggplot(overall_transition,
       aes(x = transition, y = mean_prop, fill = transition)) +
  geom_col(color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_prop - se, ymax = mean_prop + se),
                width = 0.2) +
  scale_fill_manual(values = c("Common" = "#cde5f9", "Rare" = "#ebbdc8")) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Transition",
    y = "Probability",
    title = "Overall Transition Probability"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
# rocket_transition with SE across subjects
rocket_transition <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(participant_id, choice1, transition) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(participant_id, choice1) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  group_by(choice1, transition) %>%
  summarise(
    mean_prop = mean(prop),
    se = sd(prop) / sqrt(n()),
    .groups = "drop"
  )
```

```{r, echo = FALSE}
ggplot(rocket_transition,
       aes(x = transition, y = mean_prop, fill = transition)) +
  geom_col(color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = mean_prop - se, ymax = mean_prop + se),
                width = 0.2) +
  scale_fill_manual(values = c("Common" = "#cde5f9", "Rare" = "#ebbdc8")) +
  facet_wrap(~ choice1,
             labeller = labeller(
               choice1 = c("1" = "Rocket 1", "2" = "Rocket 2")
             )) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Transition",
    y = "Probability",
    title = "Transition Probability by Rocket"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
# Recompute block_transition with SE across trials
block_transition <- stay_df_clean %>%
  mutate(transition = case_when(
    choice1 == 1 & state == 2 ~ "Common",
    choice1 == 2 & state == 3 ~ "Common",
    TRUE ~ "Rare"
  )) %>%
  group_by(participant_id) %>%
  mutate(total_trials = n()) %>%
  group_by(participant_id, transition, total_trials) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    prop = n / total_trials,
    se = sqrt(prop * (1 - prop) / total_trials)  # binomial SE for proportions
  )
```

```{r, echo = FALSE}

ggplot(block_transition,
       aes(x = transition, y = prop, fill = transition)) +
  geom_col(color = "black", width = 0.6) +
  geom_errorbar(aes(ymin = prop - se, ymax = prop + se),
                width = 0.2) +
  scale_fill_manual(values = c("Common" = "#cde5f9", "Rare" = "#ebbdc8")) +
  facet_wrap(~ participant_id, nrow = 4, ncol = 3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Transition",
    y = "Probability",
    title = "Transition Probability by Participant"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

## Log Regression - ind. level

Interpretation: 
- 7 subj with positive  β1 and negative  β3 (most subj are model based)

```{r, echo = FALSE}

indiv_models <- stay_analysis %>%
  group_by(participant_id) %>%
  nest() %>%
  mutate(
    model = map(
      data,
      ~ glm(
        stay ~ prev_reward * prev_transition,
        family = binomial,
        data = .x
      )
    ),
     results = map(model, broom::tidy)
  ) %>%
  select(participant_id, results) %>%
  unnest(results)

indiv_models

write.csv(indiv_models, file.path(output_dir, "indiv_models.csv"), row.names = FALSE)
```


```{r, echo = FALSE}
# wide format
coef_table <- stay_analysis %>%
  group_by(participant_id) %>%
  nest() %>%
  mutate(
    model = map(
      data,
      ~ glm(
        stay ~ prev_reward * prev_transition,
        family = binomial,
        data = .x
      )
    ),
    coefs = map(model, coef)
  ) %>%
  select(participant_id, coefs) %>%
  unnest_wider(coefs)

coef_table

write.csv(coef_table_indiv, file.path(output_dir, "coef_table_indiv.csv"), row.names = FALSE)
```

```{r}
# SD and CI
indiv_models %>%
  filter(term == "prev_reward:prev_transitionRare") %>%
  mutate(
    ci_lower = estimate - 1.96 * std.error,
    ci_upper = estimate + 1.96 * std.error
  ) %>%
  select(participant_id, estimate, std.error, ci_lower, ci_upper, p.value)
```


```{r}

# Convert intercept to probability:

indiv_models_conv_prob <-
  indiv_models %>%
  filter(term == "(Intercept)") %>%
  select(participant_id, estimate) %>%
  mutate(baseline_stay_prob = 1 / (1 + exp(-estimate)))

indiv_models_conv_prob

```

## Log Regression - mixed model effect

```{r, echo = FALSE}

# Fit the mixed effects model
mixed_model <- glmer(
  stay ~ prev_reward * prev_transition + 
    (1 + prev_reward * prev_transition | participant_id),
  family = binomial,
  data = stay_analysis
)

summary(mixed_model)
```
```{r, echo = FALSE}
random_effects <- broom.mixed::tidy(mixed_model, effects = "ran_vals") %>%
  select(level, term, estimate)

random_effects_wide <- broom.mixed::tidy(mixed_model, effects = "ran_vals") %>%
  select(level, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  rename(
    participant_id = level,
    intercept_re = `(Intercept)`,
    prev_reward_re = `prev_reward`,
    prev_transitionRare_re = `prev_transitionRare`,
    interaction_re = `prev_reward:prev_transitionRare`
  )

random_effects_wide

write.csv(random_effects_wide, file.path(output_dir, "random_effects_wide.csv"), row.names = FALSE)
```

## Model Fit

```{r}
Model_fit <- function(params, data) {
  
  alpha1 <- params[1]  # stage 1 learning rate
  alpha2 <- params[2]  
  lambda <- params[3]
  w  <- params[4]
  beta1  <- params[5] # inverse temp stage 1
  beta2  <- params[6]
  p      <- params[7] # perseveration
  
  # Q Tables
  Q_MF_s1 <- c(0.5, 0.5) # MF values for Stage 1 actions

  Q_MF_s2 <- matrix(0.5, nrow = 2, ncol = 2) # Q_MF_s2: MF values for Stage 2 (2 by 2 matrix all set to .5)
  
  nll      <- 0 #negative log liklihood (trying to minimize)
  prev_a1  <- NA
  
  for (t in 1:nrow(data)) {
    
    a1 <- data$choice1[t]          # 1 or 2
    a2 <- data$choice2[t]          # 1 or 2
    s2 <- data$state[t] - 1        # recode: state2→1, state3→2 (for indexing; shift down the #)
    r  <- data$won[t]              # 0 or 1
    
    # choice1==1: common→state2 (row1), rare→state3 (row2)
    # choice1==2: common→state3 (row2), rare→state2 (row1)
    Q_MB <- c(
      0.7 * max(Q_MF_s2[1, ]) + 0.3 * max(Q_MF_s2[2, ]),  # action 1
      0.3 * max(Q_MF_s2[1, ]) + 0.7 * max(Q_MF_s2[2, ])   # action 2
    )
    
    # Hybrid Q
    Q_hybrid <- (1 - w) * Q_MB + w * Q_MF_s1
    
    # Stage 1 softmax + perseveration
    rep_bonus <- c(0, 0)
    
    # assign p to whichever action was chosen on t-1
    if (!is.na(prev_a1)) rep_bonus[prev_a1] <- p
    
    logits1 <- beta1 * Q_hybrid + rep_bonus
    
    # Subtract max for numerical stability
    logits1 <- logits1 - max(logits1)
    p_a1    <- exp(logits1) / sum(exp(logits1)) #converts log into probabilities; higher log --> higher probability
    
    # Lower total NLL = model predicted subject's choices better overall
    nll     <- nll - log(p_a1[a1] + 1e-10) #prevent taking log(0)
    
    
    #  Stage 2 softmax 
    logits2 <- beta2 * Q_MF_s2[s2, ]
    logits2  <- logits2 - max(logits2)
    p_a2    <- exp(logits2) / sum(exp(logits2))
    nll     <- nll - log(p_a2[a2] + 1e-10)
    
    #  Update Q values 
    delta2 <- r - Q_MF_s2[s2, a2]   
    delta1 <- Q_MF_s2[s2, a2] - Q_MF_s1[a1]
    
    Q_MF_s2[s2, a2] <- Q_MF_s2[s2, a2] + alpha2 * delta2
    Q_MF_s1[a1]     <- Q_MF_s1[a1]     + alpha1 * lambda * delta1
    
    prev_a1 <- a1

  }
  
   return(nll)
  
}
```

```{r}
fit_subject <- function(subj_data) {
  
  best_nll    <- Inf
  best_params <- NULL
  
  for (i in 1:50) {  
    
    init <- c(
      runif(1, 0,  1),   # alpha1
      runif(1, 0,  1),   # alpha2
      runif(1, 0,  1),   # lambda
      runif(1, 0,  1),   # w 
      runif(1, 1,  8),   # beta1
      runif(1, 1,  8),   # beta2
      runif(1, -1, 1)    # p
    )
    
    result <- tryCatch(
      optim(
        par    = init,
        fn     = Model_fit,
        data   = subj_data,
        method = "L-BFGS-B",
        lower  = c(0,   0,   0,   0,   0,   0,  -5),
        upper  = c(1,   1,   1,   1,  20,  20,   5)
      ),
      error = function(e) list(value = Inf, par = rep(NA, 7))
    )
    
    if (result$value < best_nll) {
      best_nll    <- result$value
      best_params <- result$par
    }
  }
  
  data.frame(
    alpha1 = best_params[1],
    alpha2 = best_params[2],
    lambda = best_params[3],
    w      = best_params[4],
    beta1  = best_params[5],
    beta2  = best_params[6],
    p      = best_params[7],
    nll    = best_nll
  )
}
```

```{r}
# Pull one subject
test_subj <- stay_df_clean %>%
  filter(participant_id == "CCNP001339")

# Try running Model_fit directly with fixed params
test_params <- c(0.5, 0.5, 0.5, 0.5, 5, 5, 0)

Model_fit(test_params, test_subj)
```

```{r}
model_fit_results <- stay_df_clean %>%
  filter(
    choice1 %in% c(1, 2),
    choice2 %in% c(1, 2),
    won     %in% c(0, 1)
  ) %>%
  group_by(participant_id) %>%
  group_modify(~ fit_subject(.x)) %>%
  ungroup()

print(model_fit_results)
```

```{r}
# summary (model_fit_results)

model_fit_results_select <-
  model_fit_results %>%
  select(alpha1, alpha2, lambda, w, beta1, beta2, p, nll) %>%
  reframe(across(everything(), ~ quantile(.x, probs = c(0.25, 0.5, 0.75)))) %>%
  mutate(quantile = c("Q1", "Median", "Q3")) %>%
  select(quantile, everything())

model_fit_results_select
```

## Parameter Recovery

```{r}
# get a range of parameters from real fitted results

param_ranges <- model_fit_results %>%
  summarise(
    alpha1_min = min(alpha1), alpha1_max = max(alpha1),
    alpha2_min = min(alpha2), alpha2_max = max(alpha2),
    lambda_min = min(lambda), lambda_max = max(lambda),
    w_min      = min(w),      w_max      = max(w),
    beta1_min  = min(beta1),  beta1_max  = max(beta1),
    beta2_min  = min(beta2),  beta2_max  = max(beta2),
    p_min      = min(p),      p_max      = max(p)
  )


```

```{r}
# Step 2 — sample true params within real ranges
set.seed(123)
n_sim <- 10  

true_params <- data.frame(
  sim_id = 1:n_sim,
  alpha1 = runif(n_sim, param_ranges$alpha1_min, param_ranges$alpha1_max),
  alpha2 = runif(n_sim, param_ranges$alpha2_min, param_ranges$alpha2_max),
  lambda = runif(n_sim, param_ranges$lambda_min, param_ranges$lambda_max),
  w      = runif(n_sim, param_ranges$w_min,      param_ranges$w_max),
  beta1  = runif(n_sim, param_ranges$beta1_min,  param_ranges$beta1_max),
  beta2  = runif(n_sim, param_ranges$beta2_min,  param_ranges$beta2_max),
  p      = runif(n_sim, param_ranges$p_min,      param_ranges$p_max)
)

```

```{r}
# Step 3 — preallocate recovered params
recovered_params <- data.frame(
  sim_id     = 1:n_sim,
  pid        = NA_character_,
  rec_alpha1 = NA_real_,
  rec_alpha2 = NA_real_,
  rec_lambda = NA_real_,
  rec_w      = NA_real_,
  rec_beta1  = NA_real_,
  rec_beta2  = NA_real_,
  rec_p      = NA_real_
)

```

```{r}
# Step 4 — recovery loop
set.seed(123)
pids <- unique(stay_df_clean$participant_id)

for (i in 1:n_sim) {
  
  # Sample a real subject's trial structure
  sampled_pid <- sample(pids, 1)
  
  template_trials <- stay_df_clean %>%
    filter(participant_id == sampled_pid) %>%
    arrange(trial) %>%
    select(choice1, choice2, state, won)
  
  # Simulate fake choices using true params + real trial structure
  true_p <- c(
    true_params$alpha1[i],
    true_params$alpha2[i],
    true_params$lambda[i],
    true_params$w[i],
    true_params$beta1[i],
    true_params$beta2[i],
    true_params$p[i]
  )
  
  fake_data <- simulate_subject(true_p, n_trials = nrow(template_trials))
  
  # Fit model to fake choices → recover params
  recovered <- fit_subject(fake_data)
  
  recovered_params[i, 2:9] <- c(
  sampled_pid,
  recovered$alpha1, recovered$alpha2, recovered$lambda,
  recovered$w, recovered$beta1, recovered$beta2, recovered$p
)
  
  cat("\nCompleted sim", i, "of", n_sim)  # progress tracker
}

recovery_df <- cbind(true_params, recovered_params[, -1])

recovery_df <- recovery_df %>%
  mutate(across(starts_with("rec_"), as.numeric))

# Verify it worked
str(recovery_df)
```

```{r}

plot_recovery <- function(true_col, rec_col, param_name) {
  ggplot(recovery_df, aes(x = .data[[true_col]], 
                           y = .data[[rec_col]])) +
    geom_point(alpha = 0.6) +
    geom_abline(slope = 1, intercept = 0,
                color = "red", linetype = "dashed") +
    labs(x = "True", y = "Recovered", title = param_name) +
    theme_classic()
}

p1 <- plot_recovery("alpha1", "rec_alpha1", "alpha1")
p2 <- plot_recovery("alpha2", "rec_alpha2", "alpha2")
p3 <- plot_recovery("lambda", "rec_lambda", "lambda")
p4 <- plot_recovery("w",      "rec_w",      "w")
p5 <- plot_recovery("beta1",  "rec_beta1",  "beta1")
p6 <- plot_recovery("beta2",  "rec_beta2",  "beta2")
p7 <- plot_recovery("p",      "rec_p",      "p")

(p1 | p2 | p3 | p4) / (p5 | p6 | p7)
```

### Quantify the parameter recovery

```{r}
recovery_cors <- recovery_df %>%
  summarise(
    r_alpha1 = cor(alpha1, rec_alpha1, use = "complete.obs"),
    r_alpha2 = cor(alpha2, rec_alpha2, use = "complete.obs"),
    r_lambda = cor(lambda, rec_lambda, use = "complete.obs"),
    r_w      = cor(w,      rec_w,      use = "complete.obs"),
    r_beta1  = cor(beta1,  rec_beta1,  use = "complete.obs"),
    r_beta2  = cor(beta2,  rec_beta2,  use = "complete.obs"),
    r_p      = cor(p,      rec_p,      use = "complete.obs")
  )

recovery_cors
```

## Patient Info
### Demographic

```{r}
# Get participant IDs from df_all
participant_ids <- df_all$participant_id

# Filter patient_data_text by matching Q40 (participant ID column)
patient_data_text_filtered <- patient_data_text[patient_data_text$Q40 %in% participant_ids, ]

patient_data_text_selected <- patient_data_text_filtered %>%
  select(
    Q45,   # Date of Birth
    Q47,   # Gender identity
    Q50,   # Race
    Q56,   # Marital status
    Q242,  # Education
    Q64    # Employment status
  ) %>%
  mutate(
    Age            = floor(interval(mdy(Q45), Sys.Date()) / years(1)),
    Gender         = factor(Q47),
    Race           = factor(Q50),
    Marital_Status = factor(Q56),
    Education      = factor(Q242),
    Employment     = factor(Q64)
  ) %>%
  select(-Q45, -Q47, -Q50, -Q56, -Q242, -Q64) %>%
  filter(!is.na(Age), !is.na(Gender), !is.na(Race), 
         !is.na(Marital_Status), !is.na(Education), !is.na(Employment)) %>%
mutate(across(where(is.factor), droplevels))
```

```{r}
# Age row
age_row <- data.frame(
  Characteristic = "Age (years), M (SD)",
  Value = paste0(round(mean(patient_data_text_selected$Age, na.rm = TRUE), 1),
                 " (", round(sd(patient_data_text_selected$Age, na.rm = TRUE), 1), ")")
)

# Categorical rows
cat_vars <- c("Gender", "Race", "Marital_Status", "Education", "Employment")

cat_rows <- do.call(rbind, lapply(cat_vars, function(var) {
  n   <- table(patient_data_text_selected[[var]], useNA = "no")
  pct <- round(prop.table(n) * 100, 1)
  data.frame(
    Characteristic = names(n),
    Value          = paste0(as.vector(n), " (", as.vector(pct), "%)")
  )
}))

# Combine and print
demo_table <- rbind(age_row, cat_rows)
print(demo_table, row.names = FALSE)
```

### AUD severity 

```{r}
# Filter to matching participants
patient_data_num_filtered <- patient_data_num[patient_data_num$Q40 %in% participant_ids, ]

# Extract AUDIT columns
audit_data <- patient_data_num_filtered %>%
  select(Q40, Q109, Q110, Q111, Q112, Q113, Q114, Q115, Q116, Q117, Q118)

audit_data <- patient_data_num_filtered %>%
  select(Q40, Q109, Q110, Q111, Q112, Q113, Q114, Q115, Q116, Q117, Q118) %>%
  mutate(across(Q109:Q118, as.numeric)) %>%
  mutate(AUDIT_Total = rowSums(across(Q109:Q118), na.rm = TRUE))

audit_data <- audit_data[!duplicated(audit_data$Q40), ]

audit_data
```

### Random Effect & Audit

```{r}

cor1 <- cor.test(plot_data$AUDIT_Total, plot_data$prev_reward_re, method = "pearson")

label1 <- sprintf("r = %.2f, p = %.3f", cor1$estimate, cor1$p.value)

p1 <- ggplot(plot_data, aes(x = AUDIT_Total, y = prev_reward_re)) +
  geom_point(size = 3, alpha = 0.7, color = "#2196F3") +
  geom_smooth(method = "lm", se = TRUE, color = "#0D47A1", fill = "#90CAF9") +
  annotate("text", x = Inf, y = Inf, label = label1,
           hjust = 1.1, vjust = 1.5, size = 4.5) +
  labs(
    title = "Model-Free vs Alcohol Use",
    x = "AUDIT Score",
    y = "Prev_reward_re"
  ) +
  theme_classic(base_size = 13)

p1

```

```{r}
cor2 <- cor.test(plot_data$AUDIT_Total, plot_data$interaction_re, method = "pearson")

label2 <- sprintf("r = %.2f, p = %.3f", cor2$estimate, cor2$p.value)

p2 <- ggplot(plot_data, aes(x = AUDIT_Total, y = interaction_re)) +
  geom_point(size = 3, alpha = 0.7, color = "#E91E63") +
  geom_smooth(method = "lm", se = TRUE, color = "#880E4F", fill = "#F48FB1") +
  annotate("text", x = Inf, y = Inf, label = label2,
           hjust = 1.1, vjust = 1.5, size = 4.5) +
  labs(
    title = "Model-Based vs Alcohol Use",
    x = "AUDIT Score",
    y = "Interaction_re"
  ) +
  theme_classic(base_size = 13)

p2

```




